---
title: "The Good, The Bad, and The Ugle of the Beautiful Game"
author: "Debangan Dey, Andrew Pita"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(knitr)

```

# Motivation and Overview

In an era where data analysis is used to inform decision making in almost every domain of life, it is not surprising that it has taken a prominent place in the world of professional sports.  Indeed, considering that professional sports are full of quantitative data (games won, points scored, yards rushed etc), and constitute a huge international industry, sports analytics can almost be considered a sub-field of bussiness intelligence.  

Broadly, we define three areas in which data analysis can make (or already does make) a contribution to a professional sports organization: scouting for new players, optimizing game strategy, or providing real time in game recommendations. 

One well known example of using data analysis to inform scouting is the performance of MLB's Oakland Athletics in the 2002 season.  Under the direction of General Manager Billy Beane, the Athletic's organization used data analysis to develop measures that were predictive of future performance that outperformed the current measures used by other teams. In doing so, they were able to use their limited funds to build a team that went on to win the American League West title.  The strategy employed by the A's in the 2002 season has been used and further developed with great success, playing a role in the 2004 Red Sox World Series Victory, the 2016 Chicago Cubs World Series, and the 2017 Houston Astro's World Series. 

Increasingly, data analysis is being applied in the National Basketball Association and has fundamentally changed the way that coaches and managers structure their teams and approach the game.  Decades ago, teams with tall and strong post players could dominate the game (see the Lakers back to back to back championships on the back of Shaquile O'neile and Kobe Bryant).  Today, the most successful teams are built around three point shooting and speed in both ball movement and attacking.  Players that we would think of as traditional "big men" can now also shoot from the perimeter, possessing more tools than they did in the past.  This is in part due to increased athleticism and improved training on the part of the players, but also from an insight that if a player can shoot over a certain percentage from the three point line, it becomes an optimal strategy for them to take as many shots as possible. 

Finally, we return to baseball to illustrate how teams now use data analysis in real time during the game to inform strategic decision making.  Because the locations of all a batters previous successful hits can be stored digitally, constructing and updating a probability distribution of their hits during the game is an easy thing to accomplish.  Many teams now do this and shift the position of their fielders so that they can optimally defend each specific batter.  

In both professional basketball and soccer, leagues and teams have begun investing large sums of money in cameras and technology that provide not only the scoring and success percentage statistics that have always been used in the past, but also fine grained second by second information about where players and the ball are moving and how fast.  This new high resolution spatio-temporal data presents new challenges in the field of sports analytics, as teams attempt to use this data to gain a competetive advantage over other teams. 

In this project we attempt to utilize data collected by Stats Bomb during the 2018 FIFA World Cup as if we are data analysts for a soccer club, attempting to provide insights to the managers and coaching staff that can improve the teams play.  



# Related Work



# Initial Questions

We began with mostly descriptive questions to get a feel for the data and to develop more questions.  We list the specific questions below:

1) Where on the field most shots on goal came from? Do teams with different formations (the baseline positioning of the players on the field) take shots from locations at different frequencies. 

2) What patterns of play give rise to the most effective shots? A play pattern describes the setting in which events of the game unfold.  For example, if a team takes a shot that is saved and held on to by the goal keeper, then the next pattern of play begins with the keeper throwing or kicking the ball to one of his teammates.  Other examples of patterns of play include play beginning from corner kicks, from throw ins, or from set pieces following a penalty. 

3) Can we develop team level summary measures that are predictive of team performance? Data with such a fine grain of temporal detail can quickly become overwhelming. Summary measures such as these would be tremendously useful for a soccer club. 

# Data

# Exploratory Data Analysis

```{r}

load("data/Full_event_compiled.RData")

```

To start let's look into some descriptive type aspects of the data.  Specifically, let's get a feel for how many games, players, teams, shots, etc we have in this dataset. 

```{r}

total_games = length(unique(all.data$match_id))
total_players = length(unique(all.data$player.name))
total_teams = length(unique(all.data$team.name))

total_shots = nrow(unique(all.data %>% filter(type.name == "Shot") %>% 
  select(timestamp, type.name)))

total_goals = nrow(unique(all.data %>% filter(shot.outcome.name == "Goal") %>% 
  select(timestamp, shot.outcome.name)))

totals = c(total_games,total_players,total_teams, total_shots, total_goals)
names(totals) = c("Games", "Players", "Teams", "Shots", "Goals")

kable(totals,col.names = "Total")
```

So we have data from 64 games, in which 605 players took part, on 32 different teams.  The most interesting thing about this table though is the difference between the number of shots and the number of goals. Overall players are shooting at a success rate of only 11%! 

Given that goals are such a rare event, we wonder whether or not shooting percentage is associated with the number of games a team wins. Below we explore that question. 

```{r}

### seeing how many games each team won out of the total games they played

wins_frame = all.data %>% select(match_id,  home_team.home_team_name, 
                                 away_team.away_team_name, home_score, away_score)

wins_frame = wins_frame %>% group_by(home_team.home_team_name, away_team.away_team_name) %>%
  mutate(home_win = sum(home_score > away_score),
         away_win = sum(home_score < away_score)) %>% distinct

wins_frame$home_win = ifelse(wins_frame$home_win > 0, 1, 0)
wins_frame$away_win = ifelse(wins_frame$away_win > 0, 1, 0)


team_names = unique(all.data$team.name)
nloop = length(team_names)
wins_mat = matrix(nrow = nloop, ncol = 3)

for (i in 1:nloop) {
  
  team_name = team_names[i]
  team_home = wins_frame %>% filter(home_team.home_team_name == team_name)
  home_wins = sum(team_home$home_win)
  home_games = nrow(team_home)
  
  team_away = wins_frame %>% filter(away_team.away_team_name == team_name)
  away_wins = sum(team_away$away_win)
  away_games = nrow(team_home)
  
  wins = home_wins + away_wins
            
  wins_mat[i,1] = team_name
  wins_mat[i,2] = wins
  wins_mat[i,3] = home_games + away_games
  
}

```

```{r}

## getting a goal percentage for each team


```

As stated in the initial questions section, we are interested in seeing if certain play patterns give rise to more effective shots on goal.  Included in this dataset is a measure of the likelihood of a goal for any shot provided by STATSBomb called XG, or expected goal. Here we will look at differences in the mean XG and also successful goal percentage for shots taking place in different play patterns. 

```{r}

## data frame contain only shot events
shots = all.data %>% filter(type.name == "Shot") %>%
        select(match_id, timestamp, type.name, under_pressure,
                   play_pattern.name, shot.statsbomb_xg, shot.outcome.name,
                   goalkeeper.position.name, shot.deflected)

#group by the type of play pattern in which the goal took place
#mean xg for each type calculated 
play_pattern_mean_xg = shots %>% group_by(play_pattern.name) %>%
  summarise(mean_xg = mean(shot.statsbomb_xg),
            sd_xg = sd(shot.statsbomb_xg),
            n_goal = sum(shot.outcome.name == "Goal"),
            n_missed = sum(shot.outcome.name != "Goal"),
            prop_goal = n_goal/(n_missed + n_goal))

play_pattern_for_plot = shots  %>%
              select(play_pattern.name, shot.statsbomb_xg)
  
          

#now look at the same thing but simply shots on target
shots_on_target_mean_xg = shots %>% group_by(play_pattern.name) %>%
  summarise(mean_xg = mean(shot.statsbomb_xg),
            sd_xg = sd(shot.statsbomb_xg),
            n_ontarget = sum(shot.outcome.name == "Goal" | shot.outcome.name == "Blocked" | 
                           shot.outcome.name == "Saved"),
            n_total = n(),
            prop_on = n_ontarget/(n_total))

```


Now let's visualize these results. 

```{r}

ggplot(data = play_pattern_for_plot, aes(x = play_pattern.name, y = shot.statsbomb_xg)) +
  geom_boxplot() +
  labs(x = "Play Pattern", y = "Expected Goal Value") +
  theme(axis.text.x = element_text(angle = 90))

```

```{r}

ggplot(data = play_pattern_for_plot, aes(x = play_pattern.name, y = shot.statsbomb_xg)) +
  stat_summary(fun.y = mean, geom = "point") + 
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Play Pattern", y = "Expected Goal Value") 


```


# Data Analysis

# Narrative and Summary












